\textsc{Rils}-\textsc{Rols} : Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares

### 2022-11-08 paper_resources/doc/paper.tex
Update in Section 3.2

\begin{algorithm}
	\footnotesize
	\hspace*{\algorithmicindent} \textbf{Input}: input training dataset $D_{tr}$ \\
	\hspace*{\algorithmicindent} \textbf{Control parameters}: size penalty $penalty_{size}$, error tolerance $tolerance_{error}$ \\
	\hspace*{\algorithmicindent} \textbf{Output}: best symbolic formula solution $bs$
	\begin{algorithmic}[1] 
		\Procedure{\textsc{Rils}-\textsc{Rols} }{$D_{tr}$}
		\State $n_{tr} \gets |D_{tr}|$
		\State $sample_{size} \gets $ \texttt{InitialSampleSize}($n_{tr}$)
		\State $D_{tr}' \gets$ \texttt{Sample}($D_{tr}, sample_{size}$)
		\State $s \gets NodeConstant(0)$ 
		\State $s_{fit} \gets$ \texttt{Fitness}($s, D_{tr}'$)
		\State $bs, bs_{fit} \gets s, s_{fit}$ \label{line:solSet}
		\State $start_{tried}, perturbations_{tried} \gets \emptyset, \emptyset$
		\While{\emph{stopping criteria is not met}}
		\State $start_{tried} \gets start_{tried} \cup \{s\}$
		\State $s_{perturbations} \gets $ \texttt{All1Perturbations}($s$) \label{line:sPert}
		\State $s_{perturbations} \gets $ \texttt{FitOLS}($s_{perturbations}$, $D_{tr}'$)
		\State $s_{perturbations} \gets $ \texttt{OrderByR2}($s_{perturbations}$) \label{line:orderR2}
		\State $improved \gets false $
		\For{$p \in s_{perturbations}$}
		\If{$p \in perturbations_{tried}$}
		\State \textbf{continue}
		\EndIf
		\State $p \gets $ \texttt{Simplify}($p$) \label{line:simp}
		\State $p \gets $ \texttt{LocalSearch}($p$, $D_{tr}'$) \label{line:ls}
		\State $p_{fit} \gets$ \texttt{Fitness}($p$, $D_{tr}'$)
		\If{$p_{fit} < bs_{fit}$} \label{line:avoid1}
		\State $bs, bs_{fit}, improved \gets p, p_{fit}, true$ // new best solution
		\State \textbf{break}
		\EndIf \label{line:avoid2}
		\State $perturbations_{tried} \gets perturbations_{tried} \cup \{p\}$
		\EndFor
		\If{improved}
		\State $s \gets bs$ \label{line:improved}
		\Else
		\State $start_{candidates} \gets $ \texttt{All1Perturbations}($bs$)
		\If{$start_{candidates} \setminus start_{tried} = \emptyset$} // all 1-perturbations around $bs$ tried
		\State $s' \gets $ \texttt{RandomPick}($start_{candidates}$)  \label{line:pert21}
		\State $start_{candidates} \gets $ \texttt{All1Perturbations}($s'$) \label{line:pert22}
		\EndIf
		\State $s \gets $ \texttt{RandomPick}($start_{candidates} \setminus start_{tried}$) \label{line:randPick}
		\If{not improved for too many iterations}
		\State $sample_{size} \gets$ \texttt{IncreaseSampleSize}($sample_{size}, n_{tr}$) \label{line:sampleAdj1}
		\State $D_{tr}' \gets$ \texttt{Sample}($D_{tr}, sample_{size}$)\label{line:sampleAdj2}
		\EndIf
		\EndIf
		\If{$R^2$ almost 1 and RMSE almost 0 w.r.t. $tolerance_{error}$}
		\State \textbf{break} // early exit
		\EndIf
		\EndWhile
		\State $bs \gets $ \texttt{Simplify}($bs$)
		\State $bs \gets $ \texttt{RoundModelCoefficients}($bs$)
		\State \Return $bs$
		\EndProcedure
	\end{algorithmic}
	\caption{\textsc{Rils}-\textsc{Rols}  method.}
	\label{alg:rilsrols}
\end{algorithm}

@@

Now, the algorithm enters the internal loop -- it iterates over the ordered perturbations and aims to find the one which improves the best solution $bs$. But, before comparing candidate perturbation solution $p$ with $bs$, $p$ is first simplified (line~\ref{line:simp}) after which the local search is performed (line~\ref{line:ls}).
Solution simplification is done in a symbolical fashion by popular \texttt{SymPy} Python package~\cite{sympy}.
Local search tries to find local optima expressions close to the given $p$ -- explained in detail in Section~\ref{sec:ls}.  
Finally, the fitness function value of $p$ is compared to the fitness function value of $bs$. If a new incumbent is found, $bs$ is updated correspondingly and the internal loop that goes across ordered perturbations is immediately terminated (it works in a  \emph{first-improvement} strategy). Otherwise, the next perturbation is probed. 
Note that the probed perturbations are stored in a set denoted by $perturbations_{tried}$. The goal is to avoid checking the same perturbation multiple times (lines \ref{line:avoid1}-\ref{line:avoid2}), i.e. $perturbations_{tried}$ serves as a kind of tabu list, initially proposed in the known Tablu search algorithm, see~\cite{glover1998tabu}.      \fxnote{Marko: došao dovde!}

If some of $s_{perturbations}$ around starting solution $s$ yielded an improvement, $bs$ becomes the starting solution $s$ in the next iteration of the main loop (line~\ref{line:improved}). 
Otherwise, it makes no sense to set starting solution to $bs$, as nothing changed -- the search is being  \emph{trapped} in a local optimum. In order to avoid this undesired situation, a randomness needs to be triggered. First, a set of local perturbations around $bs$ is generated ($start_{candidates}$) in the same manner as before (procedure \texttt{All1Perturbations}). If at least one of these was not previously used as a starting solution ($start_{tried}$), a single perturbation from the $start_{candidates} \setminus start_{tried}$ is randomly picked (line~\ref{line:randPick}). There is a insignificant chance that $start_{candidates} \setminus start_{tried} = \emptyset$. When that happens, the set of starting solution candidates is equal to perturbations of some randomly selected perturbation of $bs$ (lines \ref{line:pert21}-\ref{line:pert22}) -- which effectively means that the perturbations with distance 2 from $bs$ are used. Indeed, there is a minor chance that these 2-perturbations will have empty set difference with $start_{tried}$ set of starting solutions. Therefore, more expensive 3-perturbations are not considered in our algorithm.

Before returning the final symbolical model, \textsc{Rils}-\textsc{Rols}  peforms the final symbolical simplification and rounding of model coefficients. The later is sensitive to control parameter $tolerance_{error}$ -- smaller value means information loss during rounding is smaller, i.e. higher number of significant digits is preserved. Note that confidence in the symbolical model is reduced when expected noise in data is higher -- the rule of thumb is: for higher expected noise $tolerance_{error}$ should be higher.

\fxnote{Dati neku smislenu procjenu kompleksnosti algoritma, ako to nije pretesko. }
\subsubsection{Fitness function}\label{sec:fitness}

Symbolic regression objective is to determine the expression governing available data. It is also allowed to obtain some equivalent expression, keeping in mind that there are multiple ways to express some symbolical equation. Although logically sound and intuitive, this objective is not quantifiable during the solution search/training phase, simply because the goal expression is not known at that point as only target values for some of the input features are known. Thus, different numerical metrics are being used in literature to guide the symbolic regression search process. The most popular are the coefficient of determination, also known as $R^2$ and mean squared error (MSE) or root mean squared error (RMSE). The important aspect of the solution quality might also be the solution expression complexity. This follows the Occam's razor principle~\cite{costa2020fast}~ that simpler solution  is more likely to be a correct one. 
The search process of \textsc{Rils}-\textsc{Rols}  is guided by the non-linear combination of $R^2$, RMSE and solution expression size (complexity) presented in Equation~\ref{eq:fitness}.

>>>>

\begin{algorithm}
	\footnotesize
	\hspace*{\algorithmicindent} \textbf{Input}: input training dataset $D_{tr}$ \\
	\hspace*{\algorithmicindent} \textbf{Control parameters}: size penalty $penalty_{size}$, error tolerance $tolerance_{error}$ \\
	\hspace*{\algorithmicindent} \textbf{Output}: best symbolic formula solution $bs$
	\begin{algorithmic}[1] 
		\Procedure{\textsc{Rils}-\textsc{Rols} }{$D_{tr}$}
		\State $n_{tr} \gets |D_{tr}|$
		\State $sample_{size} \gets $ \texttt{InitialSampleSize}($n_{tr}$)
		\State $D_{tr}' \gets$ \texttt{Sample}($D_{tr}, sample_{size}$)
		\State $s \gets NodeConstant(0)$ 
		\State $s_{fit} \gets$ \texttt{Fitness}($s, D_{tr}'$)
		\State $bs, bs_{fit} \gets s, s_{fit}$ \label{line:solSet}
		\State $start_{tried}, perturbations_{tried} \gets \emptyset, \emptyset$
		\While{\emph{stopping criteria is not met}}
		\State $start_{tried} \gets start_{tried} \cup \{s\}$
		\State $s_{perturbations} \gets $ \texttt{All1Perturbations}($s$) \label{line:sPert}
		\State $s_{perturbations} \gets $ \texttt{FitOLS}($s_{perturbations}$, $D_{tr}'$)
		\State $s_{perturbations} \gets $ \texttt{OrderByR2}($s_{perturbations}$) \label{line:orderR2}
		\State $improved \gets false $
		\For{$p \in s_{perturbations}$}
		\If{$p \in perturbations_{tried}$}
		\State \textbf{continue}
		\EndIf
		\State $p \gets $ \texttt{Simplify}($p$) \label{line:simp}
		\State $p \gets $ \texttt{LocalSearch}($p$, $D_{tr}'$) \label{line:ls}
		\State $p_{fit} \gets$ \texttt{Fitness}($p$, $D_{tr}'$)
		\If{$p_{fit} < bs_{fit}$} \label{line:avoid1}
		\State $bs, bs_{fit}, improved \gets p, p_{fit}, true$ // new best solution
		\State \textbf{break}
		\EndIf \label{line:avoid2}
		\State $perturbations_{tried} \gets perturbations_{tried} \cup \{p\}$
		\EndFor
		\If{improved}
		\State $s \gets bs$ \label{line:improved}
		\Else
		\State $start_{candidates} \gets $ \texttt{All1Perturbations}($bs$)
		\If{$start_{candidates} \setminus start_{tried} = \emptyset$} // all 1-perturbations around $bs$ tried
		\State $s' \gets $ \texttt{RandomPick}($start_{candidates}$)  \label{line:pert21}
		\State $start_{candidates} \gets $ \texttt{All1Perturbations}($s'$) \label{line:pert22} // 2-permutations of \textit{bs}
		\EndIf
		\State $s \gets $ \texttt{RandomPick}($start_{candidates} \setminus start_{tried}$) \label{line:randPick}
		\If{not improved for too many iterations}
		\State $sample_{size} \gets$ \texttt{IncreaseSampleSize}($sample_{size}, n_{tr}$) \label{line:sampleAdj1}
		\State $D_{tr}' \gets$ \texttt{Sample}($D_{tr}, sample_{size}$)\label{line:sampleAdj2}
		\EndIf
		\EndIf
		\If{$R^2$ almost 1 and RMSE almost 0 w.r.t. $tolerance_{error}$}
		\State \textbf{break} // early exit
		\EndIf
		\EndWhile
		\State $bs \gets $ \texttt{Simplify}($bs$)
		\State $bs \gets $ \texttt{RoundModelCoefficients}($bs$)
		\State \Return $bs$
		\EndProcedure
	\end{algorithmic}
	\caption{\textsc{Rils}-\textsc{Rols}  method.}
	\label{alg:rilsrols}
\end{algorithm}

@@

Now, the algorithm enters the internal loop -- it iterates over the ordered perturbations and aims to find the one which improves the best solution $bs$. But, before comparing candidate perturbation solution $p$ with $bs$, $p$ is first simplified (line~\ref{line:simp}) after which the local search is performed (line~\ref{line:ls}).
Solution simplification is done in a symbolical fashion by popular \texttt{SymPy} Python package~\cite{sympy}.
Local search tries to find local optima expressions close to the given $p$ -- explained in detail in Section~\ref{sec:ls}.  
Finally, the fitness function value of $p$ is compared to the fitness function value of $bs$. If a new incumbent is found, $bs$ is updated correspondingly and the internal loop that goes across ordered perturbations is immediately terminated (it works in a  \emph{first-improvement} strategy). Otherwise, the next perturbation is probed. 
Note that the probed perturbations are stored in a set denoted by $perturbations_{tried}$. The goal is to avoid checking the same perturbation multiple times (lines \ref{line:avoid1}-\ref{line:avoid2}), i.e. $perturbations_{tried}$ serves as a kind of tabu list, which is known from the Tablu search meta-heuristic, see~\cite{glover1998tabu}.

If some of $s_{perturbations}$ around starting solution $s$ yielded an improvement, $bs$ becomes the starting solution $s$ in the next iteration of the main loop (line~\ref{line:improved}). 
Otherwise, it makes no sense to set starting solution to $bs$, as nothing changed -- the search is being  \emph{trapped} in a local optimum. In order to avoid this undesired situation, randomness needs to be triggered. First, a set of local perturbations around $bs$ is generated ($start_{candidates}$) in the same manner as before (procedure \texttt{All1Perturbations}). If at least one of these was not previously used as a starting solution ($start_{tried}$), a single perturbation from the $start_{candidates} \setminus start_{tried}$ is randomly picked (line~\ref{line:randPick}). There is a minor chance that $start_{candidates} \setminus start_{tried} = \emptyset$. When that happens, the set of starting solution candidates is equal to perturbations of some randomly selected perturbation of $bs$ (lines \ref{line:pert21}-\ref{line:pert22}) -- which effectively means that the perturbations with distance 2 from $bs$ are used. Indeed, there is almost no chance that these 2-perturbations will have an empty set difference with $start_{tried}$ set of starting solutions. Therefore, more expensive 3-perturbations will not be considered in our algorithm.

Before returning the final symbolical model, \textsc{Rils}-\textsc{Rols}  performs the final symbolical simplification and rounding of model coefficients. The latter is sensitive to control parameter $tolerance_{error}$ -- a smaller value means the information loss during rounding is smaller, i.e. a higher number of significant digits is preserved. Note that confidence in the symbolical model is reduced when expected noise in data is higher -- the rule of thumb is: for higher expected noise $tolerance_{error}$ should be higher.

\fxnote{Dati neku smislenu procjenu kompleksnosti algoritma, ako to nije pretesko. }
\subsubsection{Fitness function}\label{sec:fitness}
\fxnote{Marko: stigao do ovog dijela...}
Symbolic regression objective is to determine the expression governing available data. It is also allowed to obtain some equivalent expression, keeping in mind that there are multiple ways to express some symbolical equation. Although logically sound and intuitive, this objective is not quantifiable during the solution search/training phase, simply because the goal expression is not known at that point as only target values for some of the input features are known. Thus, different numerical metrics are being used in literature to guide the symbolic regression search process. The most popular are the coefficient of determination, also known as $R^2$ and mean squared error (MSE) or root mean squared error (RMSE). The important aspect of the solution quality might also be the solution expression complexity. This follows the Occam's razor principle~\cite{costa2020fast}~ that simpler solution  is more likely to be a correct one. 
The search process of \textsc{Rils}-\textsc{Rols}  is guided by the non-linear combination of $R^2$, RMSE and solution expression size (complexity) presented in Equation~\ref{eq:fitness}.
